---
title: "Final Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, echo=FALSE}
library(prettydoc)
library(data.table)
library(plyr)
library(Hmisc)
library(DT)
library(imputeTS)
library(VIM)
#library(rworldmap)
```

## Introduction


```{r read_data, echo=FALSE}

dat <- fread(input = "../final project/metadata/data.csv", verbose = FALSE)
region<-fread(input = "../final project/metadata/Metadata_Country_API_SH.HIV.INCD.ZS_DS2_en_csv_v2.csv", verbose = FALSE)
category<- fread(input = "../final project/metadata/Category.csv", verbose = FALSE)

datatable(dat[1:10,])

```

## Data cleaning

The main issue with the orignal data set is the structure and structural missing data. The columns of the orignal data are time series data and each row represents a specific indicator of a country. This structure is the transpose of the structure we usually use in R.So the first step of data cleaning is to adjust this dataset into a tidy dataset.

```{r data_melt}
##Remove the space in column name
old_name<-names(dat)[1:4]
new_name<-gsub(pattern = " ",replacement = ".",x=old_name)
setnames(x=dat,old = old_name,new = new_name)
##Stroe the indicator name and code into a new table and remove the indicator name from the original dataset.
Indicator<-dat[,3:4]
Indicator<-Indicator[,.(Indicator.Code=unique(Indicator.Code)),by=Indicator.Name]
sum(dat[,!is.na(V61)])
print("V61 is an emplty column")
dat[,`:=`(Indicator.Name=NULL,V61=NULL)]
##Reconstruct the data 
id.vars<-names(dat)[1:3]
measure.vars<-as.character(1960:2015)
mdat<-melt(data = dat,id.vars=1:3, measure.vars=measure.vars, variable.name = "year",variable.factor = FALSE)
mdat[, `:=`(value, mapvalues(x = value,from = "", to = NA))]
mdat[,`:=`(year=as.numeric(year),value=as.numeric(value))]
dcast_formula<-"Country.Code+Country.Name+year~Indicator.Code"
mdat<-dcast(mdat[],dcast_formula)
data.table(mdat[1:10, 1:5])

```

Since the original data set is a humongous dataset (348 columns in total), we want to pick up the variables that are only related to our analysis. To make futher analysis easier, we will classify each indicator and group the indicator that we are going to use in following analysis. To achieve this, first we manually picked 84
```{r subset}
##Subset the indicator we want and the data from 1990-2015
data.table(category[1:10, ])
sub_mdat_pre<-mdat[year%in%1990:2015,.SD,.SDcols=c("Country.Code","Country.Name","year",unique(category$`Indicator.Code`))]
##Use world bank Region Codelist to catefory data by region and subset only the region we want
region<-region[, 1:3]
sub_mdat_pre<-merge(x=sub_mdat_pre,y=region[,],by.y="Country Code",by.x="Country.Code",all.x=TRUE)
sub_region<-c("Latin America & Caribbean","South Asia","Sub-Saharan Africa","Middle East & North Africa")
sub_mdat_pre<-sub_mdat_pre[Region%in%sub_region,]
data.table(sub_mdat_pre[])
```


```{r data_cleaning}
mean.missing <- function(x) {
return(mean(is.na(x)))
}
# calcualte the missing rate for each column (variable) by year
missing.rate <-sub_mdat_pre[,lapply(.SD, mean.missing),by=year]

# evaluate the avarage missing rate for each column
average.missing.rate <- missing.rate[, lapply(.SD, FUN = mean)]



# the length of average missing rate less than 80%
cat("the number of variables that missing rate more than 20% is", length(which(average.missing.rate < 0.2)))

cat("\n the number of current countries is", length(unique(sub_mdat_pre$Country.Name)))
##first we want to sub set the country which have the record of Prevalence of HIV



##Data for linear regression
HIV.prevalence<-c("SH.HIV.1524.FE.ZS","SH.HIV.1524.MA.ZS","SH.DYN.AIDS.ZS")
HIV.prevalence.missing<-sub_mdat_pre[,lapply(.SD,mean.missing),by=Country.Name,.SDcols=HIV.prevalence]
HIV.prevalence.missing[,sum.missing:=rowSums(HIV.prevalence.missing[,-1])]

sub.country<-HIV.prevalence.missing[sum.missing==0,Country.Name]
sub_mdat_pre<- sub_mdat_pre[Country.Name%in%sub.country,]
  
missing_year<-sub_mdat_pre[,lapply(.SD,mean.missing),by=year]

# We calclate the missing value rate by year and figure out how many years have missing value rate over 0.7 for each indicator.
number.missing.years <- missing_year[, lapply(.SD, function(x) sum(x>0.7))]
# Next we calculate the total number of years that each indicator have missing value over 0.7. If a indicator has more than 24 years of missing value rate over 0.7, we will put it aside for now 
chosen.variables.by.year <-names(missing_year)[which(number.missing.years<22)]
sub.mdat<-sub_mdat_pre[,.SD,.SDcols=c("year",chosen.variables.by.year)]


##export the data for time serise ploting
sub.mdat.plot.ts<-sub_mdat_pre[,]
write.csv(x = sub.mdat.plot.ts,file = "../final project/metadata/ploting data.csv")
total.missing.rate<-sub.mdat[,lapply(.SD,mean.missing),by=year]

#Next we want to study the pattern of missing value rate and find out which is structural missing and which is random missing 
total.missing.rate[,]
#This function will return the number of different missing value an indicator have(except 1)
missing.type<-function(x){
  missing<-unique(x)
  num.missing<-sum(missing!=1)
  return(num.missing)
}
num.missing<-total.missing.rate[,lapply(.SD, missing.type)]

# for indicators which have unique missing  value from beginning, we consider is structural missing and do not need futher process 

structural.missing<-c("year",names(num.missing)[which(num.missing==1)])
id.var<-c("year", "Country.Code","Country.Name", "Region","IncomeGroup")
total.missing.rate[,.SD,.SDcols=c(id.var,names(num.missing)[names(num.missing)%nin%structural.missing])]


##impute 

num.missing <- function(x) {
return(sum(!is.na(x)))
}

##检查一开始全部missing
structural<-function(x){
  lengths<-rle(x)$lengths
  values<-rle(x)$values
  if (values[1]==1){
    length.structural<-lengths[1]
  }
  else{
    length.structural<-0
  }
  return(length.structural)
  
}

#num.missing
num.structural<-total.missing.rate[,lapply(.SD,structural)]
#time serise interpolation
structural.imputation <- function(x,missing.number) {
  missing.rate<-sum(is.na(x))/(length(x)-missing.number)
  if(missing.rate>0.25){
    return(x)
  }
  else{
    if(missing.number==0){
      x<-na.interpolation(x=x,option = "spline",method="natural")
      return(x)
    }
    else{
      x[-(1:missing.number)]<-na.interpolation(x=x[-(1:missing.number)],option = "spline",method="natural")
      return(x)
    }
}
}

sub.mdat[,lapply(.SD,structural.imputation,missing.number=0),by=Country.Name]
sub.mdat[,index:=1:.N]
sub.impute.dat<-sub.mdat[,.(index)]

for (i in 1:length(unique(t(num.structural)))) {
  num<-unique(t(num.structural))[i]
  contain.structural<-names(num.structural)[which(t(num.structural)==num)]
  temp.dat<-sub.mdat[,lapply(.SD,structural.imputation,missing.number=num),by=Country.Name,.SDcols=c("index",contain.structural)]
  temp.dat[,`:=`(Country.Name=NULL)]
  sub.impute.dat<-merge(sub.impute.dat,temp.dat,by="index")
}

sub.impute.dat[,`:=`(index=NULL)]
sub.mdat[,`:=`(index=NULL)]
##sub.impute.dat isimputed datatable

##-------Show resul---------------
sub.impute.dat[,lapply(.SD,mean.missing),]
sub.mdat[,lapply(.SD,mean.missing),.SDcols=names(sub.impute.dat)[-1]]
##--------check difference after impute----------------
#unify dimmension
the.summary<-function(x){
  if (sum(is.na(x))==0){
    the.summary<-summary(x)
    return(c(the.summary,"NA's"=0))
  }
  else{
    the.summary<-summary(x)
    return(the.summary)
  }
}

value.var<-names(sub.mdat)[names(sub.mdat)%nin%id.var]
impute.after<-sub.impute.dat[,lapply(.SD,the.summary),.SDcols=value.var]
impute.before<-sub.mdat[,lapply(.SD,the.summary),.SDcols=value.var]

diff.impute<-(impute.after[,lapply(.SD, as.numeric)]-impute.before[,lapply(.SD,as.numeric)])/impute.before[,lapply(.SD, as.numeric)]
parameter<-names(sub.mdat[,summary(NY.GNP.PCAP.CD)])
diff.impute[,patameter:=parameter]
diff.impute[,.SD,by="parameter"]
```
From code above, we narrow down 71 variables to 49. We'll do more investigation of these these 49 variable next week.
```{r data_merging}
# region<-as.data.table(countryRegions[,c("ISO3","REGION")])
# mdat<-merge(x=mdat,y=region,by.y="ISO3",by.x="Country.Code",all.x=TRUE)
#grep(pattern = "AIDS",x=names(sub.mdat),value = TRUE)

a<-function(x){
  missing<-which(is.na(x))
  if(length(missing)!=0){
    x[missing]<-median(x,na.rm=TRUE)
  }
  return(x)
}
sub.mdat[,a(NY.GNP.PCAP.CD),by=c("year","IncomeGroup")]
sub.mdat[,lapply(.SD, a),by=c("year","IncomeGroup")]
```
length(unique(t(num.structural)))

