---
title: "Team Bazinga! RMarkdown Part"
author: "Ni Zhan, Yiyuan Wang, Liang Wei"
date: "12/15/2017"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, eval = TRUE, comment="", warning = FALSE, message = FALSE, tidy.opts=list(width.cutoff=55), tidy = TRUE)
```


```{r libraries, echo=FALSE}
library(prettydoc)
library(data.table)
library(plyr)
library(Hmisc)
library(DT)
library(imputeTS)
library(reshape2)
```

## Data Loading


```{r read_data, echo=FALSE}
#================================Load Data===========================================
dat <- fread(input = "../data/data.csv", verbose = FALSE)
region<-fread(input = "../data/Metadata_Country_API_SH.HIV.INCD.ZS_DS2_en_csv_v2.csv", verbose = FALSE)
category<- fread(input = "../data/Category.csv", verbose = FALSE)


```

## Data Examination

The main issue with the orignal data set is the structure and structural missing data. The columns of the orignal data are time series data and each row represents a specific indicator of a country. This structure is the transpose of the structure we usually use in R.So the first step of data cleaning is to adjust this dataset into a tidy dataset.

```{r data_melt}
#================================Data Reshape=========================
##Remove the space in column name
old_name<-names(dat)[1:4]
new_name<-gsub(pattern = " ",replacement = ".",x=old_name)
setnames(x=dat,old = old_name,new = new_name)

##Stroe the indicator name and code into a new table and remove the indicator name from the original dataset.
Indicator<-dat[,3:4]
Indicator<-Indicator[,.(Indicator.Code=unique(Indicator.Code)),by=Indicator.Name]
sum(dat[,!is.na(V61)])
#print("V61 is an emplty column")
dat[,`:=`(Indicator.Name=NULL,V61=NULL)]

##Reconstruct the data 
id.vars<-names(dat)[1:3]
measure.vars<-as.character(1960:2015)
mdat<-melt(data = dat,id.vars=1:3, measure.vars=measure.vars, variable.name = "year",variable.factor = FALSE)
mdat[, `:=`(value, mapvalues(x = value,from = "", to = NA))]
mdat[,`:=`(year=as.numeric(year),value=as.numeric(value))]
dcast_formula<-"Country.Code+Country.Name+year~Indicator.Code"
mdat<-dcast(mdat[],dcast_formula)

#data.table(mdat)
setDT(mdat)
#class(mdat)
cat("Original missing values are", sum(is.na(mdat)))
cat("\nOriginal missing rate is", round(sum(is.na(mdat)) / (14448*348) * 100, digits = 2), "%")
```


## Data Range Selection & Adding Region Variable
```{r subset}
##Subset the indicator we want and the data from 1990-2015
sub_mdat_pre<-mdat[year%in%1990:2015,.SD,.SDcols=c("Country.Code","Country.Name","year",unique(category$`Indicator.Code`))]

##Use world bank Region Codelist to catefory data by region and subset only the region we want
region<-region[, 1:3]
sub_mdat_pre<-merge(x=sub_mdat_pre,y=region[,],by.y="Country Code",by.x="Country.Code",all.x=TRUE)
sub_region<-c("Latin America & Caribbean","South Asia","Sub-Saharan Africa","Middle East & North Africa")
sub_mdat_pre<-sub_mdat_pre[Region%in%sub_region,]

cat("modified number of countries according to region:", length(unique(sub_mdat_pre$Country.Code)))
```


## Data Cleaning
```{r data_cleaning}
mean.missing <- function(x) {
return(mean(is.na(x)))
}

# calcualte the missing rate for each column (variable) by year
missing.rate <-sub_mdat_pre[,lapply(.SD, mean.missing),by=year]

# evaluate the avarage missing rate for each column
average.missing.rate <- missing.rate[, lapply(.SD, FUN = mean)]

##Data for linear regression
HIV.prevalence<-c("SH.HIV.1524.FE.ZS","SH.HIV.1524.MA.ZS","SH.DYN.AIDS.ZS")
HIV.prevalence.missing<-sub_mdat_pre[,lapply(.SD,mean.missing),by=Country.Name,.SDcols=HIV.prevalence]
HIV.prevalence.missing[,sum.missing:=rowSums(HIV.prevalence.missing[,-1])]

sub.country<-HIV.prevalence.missing[sum.missing==0,Country.Name]
sub_mdat_pre<- sub_mdat_pre[Country.Name%in%sub.country,]
  
missing_year<-sub_mdat_pre[,lapply(.SD,mean.missing),by=year]
total.missing <- mdat[, lapply(.SD,FUN=mean.missing)]

# We calclate the missing value rate by year and figure out how many years have missing value rate over 0.7 for each indicator.
number.missing.years <- missing_year[, lapply(.SD, function(x) sum(x>0.7))]

# Next we calculate the total number of years that each indicator have missing value over 0.7. If a indicator has more than 22 years of missing value rate over 0.7, we will put it aside for now 
chosen.variables.by.year <-names(missing_year)[which(number.missing.years<22)]
sub.mdat<-sub_mdat_pre[,.SD,.SDcols=c("year",chosen.variables.by.year)]
cat("The number of variables after choosing by mean missing rate less than 70% and variables exist at least more than 3 years is", length(unique(names(sub.mdat))))

##export the data before imputing
sub.mdat.plot.ts<-sub_mdat_pre[,]

cat("\nThe number of countries before imputation is", length(unique(sub_mdat_pre$Country.Code)))
#write.csv(x = sub.mdat, "~/Downloads/before_imputation.csv")

total.missing.rate<-sub.mdat[,lapply(.SD,mean.missing),by=year]

#Next we want to study the pattern of missing value rate and find out which is structural missing and which is random missing 
#total.missing.rate[,]

#This function will return the number of different missing value an indicator have(except 1)
missing.type<-function(x){
  missing<-unique(x)
  num.missing<-sum(missing!=1)
  return(num.missing)
}

num.missing<-total.missing.rate[,lapply(.SD, missing.type)]

# for indicators which have unique missing  value from beginning, we consider is structural missing and do not need futher process 

structural.missing<-c("year",names(num.missing)[which(num.missing==1)])
id.var<-c("year", "Country.Code","Country.Name", "Region","IncomeGroup")
#total.missing.rate[,.SD,.SDcols=c(id.var,names(num.missing)[names(num.missing)%nin%structural.missing])]
```


## Linear Analysis for the data before imputation.
We were curious about the bias/difference between before and after imputation
```{r}
#===================================Linear Regression Model==========================
round.numerics <- function(x, digits = 0, nearest = 1){
  if(is.numeric(x)){
    return(nearest * round(x = x/nearest, digits = digits))
  }
  else{
    return(x)
  }
}

logistic.regression.summary <- function(glm.mod, digits = 3) {
  glm.coefs <- as.data.table(summary(glm.mod)$coefficients, keep.rownames = TRUE)
  alpha = 0.05
  z <- qnorm(p = 1 - alpha/2, mean = 0, sd = 1)
  glm.coefs[, `:=`(Odds.Ratio, exp(Estimate))]
  glm.coefs[, `:=`(OR.Lower.95, exp(Estimate - z * `Std. Error`))]
  glm.coefs[, `:=`(OR.Upper.95, exp(Estimate + z * `Std. Error`))]
  return(glm.coefs[])
}

linear.regression.summary <- function(lm.mod, digits = 3) {
  lm.coefs <- as.data.table(summary(lm.mod)$coefficients, keep.rownames = TRUE)
  alpha = 0.05
  z <- qnorm(p = 1 - alpha/2, mean = 0, sd = 1)
  lm.coefs[, `:=`(Odds.Ratio, exp(Estimate))]
  lm.coefs[, `:=`(Coef.Lower.95, exp(Estimate - z * `Std. Error`))]
  lm.coefs[, `:=`(Coef.Upper.95, exp(Estimate + z * `Std. Error`))]
  return(lm.coefs)
}


fit.model <- function(dat, the.formula, digits = 3) {
  dat <- as.data.table(dat)
  the.outcome <- trimws(strsplit(x = the.formula, split = "~", fixed = TRUE)[[1]][1])
  unique.values <- dat[, unique(get(the.outcome))]
  unique.values <- unique.values[!is.na(unique.values)]
  if (length(unique.values) == 2) {
    mod <- glm(formula = the.formula, family = "binomial", data = dat)
    mod.summary <- logistic.regression.summary(glm.mod = mod, digits = digits)
  } else {
    mod <- lm(formula = the.formula, data = dat)
    mod.summary <- linear.regression.summary(lm.mod = mod, digits = digits)
  }
  mod.summary.rounded <- mod.summary[, lapply(X = .SD, FUN = "round.numerics", digits = digits)]
  return(mod.summary.rounded)
}

#===================Linear Regression Analysis Before Imputation=====================
hiv.var <- category[Group == "HIV"]$Indicator.Code
cols.not.hiv <- names(sub.mdat)[names(sub.mdat) %nin% hiv.var]
cols.in.formula <- cols.not.hiv[cols.not.hiv %nin% c("V1", "year", "Country.Code", "Country.Name")]
cols.in.formula <- paste(cols.in.formula, collapse = "+")

the.formula <- paste("SH.DYN.AIDS.ZS", "~", cols.in.formula)
hiv.15.49.total <- fit.model(dat = sub.mdat, the.formula = the.formula)
datatable(data = hiv.15.49.total)
```

## Impute by Time Series
```{r}
##=================================Imputing by Time Series============================
num.missing <- function(x) {
return(sum(!is.na(x)))
}


##checking missing
structural<-function(x){
  lengths<-rle(x)$lengths
  values<-rle(x)$values
  if (values[1]==1){
    length.structural<-lengths[1]
  }
  else{
    length.structural<-0
  }
  return(length.structural)
  
}

#num.missing
num.structural<-total.missing.rate[,lapply(.SD,structural)]
#time serise interpolation
structural.imputation <- function(x,missing.number) {
  missing.rate<-sum(is.na(x))/(length(x)-missing.number)
  if(missing.rate>0.25){
    return(x)
  }else{
    station.test <- Box.test(na.remove(x), lag = 2)
    # test stationary, if not, don't impute, o/w, use ts to impute
    if(station.test$p.value > 0.05 | is.nan(station.test$p.value)){
    return(x)
  }
  else{
    if(missing.number == 0){
      x <- na.interpolation(x=x,option = "spline",method="natural")
      return(x)
    } else{
        x[-(1:missing.number)]<-na.interpolation(x=x[-(1:missing.number)],option = "spline",method="natural")
        return(x)
    }
  }
  }
}

#sub.mdat[,lapply(.SD,structural.imputation,missing.number=0),by=Country.Name, .SDcols = "NY.GNP.PCAP.CD"]
sub.mdat[,index:=1:.N]
sub.impute.dat<-sub.mdat[,.(index)]

for (i in 1:4) {
  num<-unique(t(num.structural))[i]
  
  contain.structural<-names(num.structural)[which(t(num.structural)==num)]
  
  temp.dat<-sub.mdat[,lapply(.SD,structural.imputation,missing.number=num),by=Country.Name,.SDcols=c("index",contain.structural[contain.structural %nin% id.var])]
  temp.dat[,`:=`(Country.Name=NULL)]
  sub.impute.dat<-merge(sub.impute.dat,temp.dat,by="index")
}

sub.impute.dat<-merge(sub.mdat[,.SD,.SDcols=c("index",id.var)],sub.impute.dat,by="index")
sub.impute.dat[,`:=`(index=NULL)]
sub.mdat[,`:=`(index=NULL)]
##sub.impute.dat isimputed datatable

##-------Show resul---------------
#sub.impute.dat[,lapply(.SD,mean.missing),]
#sub.mdat[,lapply(.SD,mean.missing),.SDcols=names(sub.impute.dat)[]]
##--------check difference after impute----------------
#unify dimmension
the.summary<-function(x){
  if (sum(is.na(x))==0){
    the.summary<-summary(x)
    return(c(the.summary,"NA's"=0))
  }
  else{
    the.summary<-summary(x)
    return(the.summary)
  }
}

value.var<-names(sub.mdat)[names(sub.mdat)%nin%id.var]
impute.after<-sub.impute.dat[,lapply(.SD,the.summary),.SDcols=value.var]
impute.before<-sub.mdat[,lapply(.SD,the.summary),.SDcols=value.var]

diff.impute<-(impute.after[,lapply(.SD, as.numeric)]-impute.before[,lapply(.SD,as.numeric)])/impute.before[,lapply(.SD, as.numeric)]
parameter<-names(sub.mdat[,summary(NY.GNP.PCAP.CD)])
#diff.impute[,patameter:=parameter]
#diff.impute[,.SD,by="parameter"]

cat("after time series imputation, the number of missing valuse is:", sum(is.na(sub.impute.dat)))
```


## Impute by Income Group
```{r data_merging}
# region<-as.data.table(countryRegions[,c("ISO3","REGION")])
# mdat<-merge(x=mdat,y=region,by.y="ISO3",by.x="Country.Code",all.x=TRUE)
#grep(pattern = "AIDS",x=names(sub.mdat),value = TRUE)

#===================================Impute by IncomeGroup function===============================
imputation.by.income <- function(x){
  missing <- which(is.na(x))
  if(length(missing) != 0){
    x[missing] <- as.double(median(x, na.rm = TRUE))
  }
  return(x)
}

#=================================Start Imputing by IncomeGroup after time series=================
# calculate current missing rate
# build an index column for convenience
sub.impute.dat$V1 <- seq.int(nrow(sub.impute.dat))
#sub.impute.dat[, "V1"] <- sapply(sub.impute.dat[, "V1"], as.numeric)
after.ts.missing.rate <- sub.impute.dat[, lapply(.SD, FUN = mean.missing)]
missing.not.zero <- names(after.ts.missing.rate)[which(after.ts.missing.rate != 0)]

#==================Finance Group Imputing by IncomeGroup=============
income.related.colname <- category[Group == "Finance" & Indicator.Code %in% missing.not.zero]$Indicator.Code
income.related.cols <- c("V1", income.related.colname)
imputed.by.income.group <- sub.impute.dat[, lapply(.SD, imputation.by.income), .SDcols = income.related.cols, by = c("year", "IncomeGroup")]

setorderv(imputed.by.income.group, "V1")

#=======comparison========
# the result is the missing rate decreased by impution of using income group
#sub.impute.dat[, lapply(.SD, mean.missing), .SDcols = income.related.cols] - imputed.by.income.group[, lapply(.SD, FUN = mean.missing)][, -c(1,2)]

#=======merging===========
result.missing.after.income <- imputed.by.income.group[, lapply(.SD, FUN = mean.missing)]
# from above, since the missing rate is same, it's reasonable to believe that they are structural missing, and it's confirmed by inspection

# merge 
after.finance.income <- merge(sub.impute.dat[, !(income.related.colname), with = FALSE],  imputed.by.income.group[, -c(1,2)], by = "V1")


#====================================Eduction Group Imputing by IncomeGroup=======================
education.related.colnames <- category[Group == "Education" & Indicator.Code %in% missing.not.zero]$Indicator.Code
education.related.cols <- c("V1", education.related.colnames)
education.imputed.by.income <- sub.impute.dat[, lapply(.SD, imputation.by.income), .SDcols = education.related.cols, by = c("year", "IncomeGroup")]
setorderv(education.imputed.by.income, "V1")

# ========comparision======
#sub.impute.dat[, lapply(.SD, mean.missing), .SDcols = education.related.cols] - education.imputed.by.income[, lapply(.SD, FUN = mean.missing)][, -c(1,2)]

#=========merging===========
after.edu.finance <- merge(sub.impute.dat[, !(education.related.colnames), with = FALSE],  education.imputed.by.income[, -c(1,2)], by = "V1")



#===================================Health Service Imputing by IncomeGroup=======================
health.related.colnames <- category[Group == "Health service" & Indicator.Code %in% missing.not.zero]$Indicator.Code
health.related.cols <- c("V1", health.related.colnames)
health.imputed.by.income <- sub.impute.dat[, lapply(.SD, imputation.by.income), .SDcols = health.related.cols, by = c("year", "IncomeGroup")]
setorderv(education.imputed.by.income, "V1")

#=======comparision=========
#sub.impute.dat[, lapply(.SD, mean.missing), .SDcols = health.related.cols] - health.imputed.by.income[, lapply(.SD, FUN = mean.missing)][, -c(1,2)]

#=======merging=============
after.edu.finance.health <- merge(sub.impute.dat[, !(education.related.colnames), with = FALSE],  education.imputed.by.income[, -c(1,2)], by = "V1")

after.edu.finance.health$V1 <- NULL
sub.impute.dat$V1 <- NULL
after.imputation <- after.edu.finance.health
#write.csv(x = after.imputation, "~/Downloads/after_imputation.csv")
cat("the final missing rate after imputation is:", round(sum(is.na(after.imputation)) / 101088 *100, digits = 2),"%")
```


## Adding Population Variable to Standarlize HIV variables
```{r adding variable}
#===============================Adding Population===========================================
current <- fread(input = "../data/after_imputation.csv")
before <- fread(input = "../data/before_imputation.csv")
population <- fread(input = "../data/population.csv")

current.pop <- population[`Country Name` %in% current$Country.Name]
chosen.year<-c("Country Name",as.character(1990:2015))
pop.table <- current.pop[, .SD,.SDcols = chosen.year]

names(pop.table)[1] <- "Country.Name"
#which(is.na(pop.table))

pop.reshape <- melt(data = pop.table, id.vars = "Country.Name", value.name = "SP.POP.TOTL")
names(pop.reshape)[2] <- "year"
pop.reshape[,year:=as.integer(as.character(year))]
with.population <- merge(after.imputation, pop.reshape, by = c("Country.Name", "year"),all=TRUE)

#==================divided hiv stats by population
population.percent<- c("SH.MED.BEDS.ZS","SH.MED.NUMW.P3","SH.MED.PHYS.ZS","SH.HIV.TOTL","SH.HIV.INCD.TL","SH.DYN.AIDS","SH.HIV.INCD","SH.DYN.AIDS.DH","SH.HIV.0014","SH.HIV.INCD.14","SH.HIV.ORPH","SH.DTH.IMRT","SH.MMR.DTHS","SH.DTH.NMRT")
with.population[,SP.POP.TOTL:=as.numeric(SP.POP.TOTL)]
hiv.percent <- with.population[,.SD, .SDcols = population.percent]/with.population$SP.POP.TOTL*1000
 
with.population.percentage <- cbind(with.population[, !(population.percent), with = FALSE],  hiv.percent)
#with.population.percentage[, `:=`(V1, NULL)]
#write.csv(with.population.percentage, "~/Downloads/final_dat.csv")

#==================================Adding population info for before_imputation data
before.with.pop <- merge(before, pop.reshape, by = c("Country.Name", "year"), all = TRUE)
before.with.pop[,SP.POP.TOTL:=as.numeric(SP.POP.TOTL)]
before.hiv.percent <- before.with.pop[,.SD, .SDcols = population.percent]/with.population$SP.POP.TOTL*1000

before.with.pop.per <- cbind(before.with.pop[, !(population.percent), with = FALSE], before.hiv.percent)
before.with.pop.per[, `:=`(V1, NULL)]
#write.csv(before.with.pop.per, "~/Downloads/final_before.csv")
```


## Graphs
```{r plot}
#==================================plot hiv info world map=====================
library(rworldmap)
library(RColorBrewer)
library(VIM)
library(corrplot)
library(qgraph)
subdat <- current[, c("year","Country.Code", "Country.Name", "SH.DYN.AIDS.ZS")]

reshaped.dat <- dcast(subdat, Country.Name ~ year, value.var = "SH.DYN.AIDS.ZS")

match.country <- joinCountryData2Map(reshaped.dat, joinCode="NAME", nameJoinColumn="Country.Name")

colourPalette <- brewer.pal(8, 'RdPu')

mapCountryData(match.country, nameColumnToPlot="2015", missingCountryCol = gray(.8), colourPalette = colourPalette, catMethod = c(0,5,10,15,20,25,30))

#===================================plot missing======================
#missing_plot <- aggr(mdat, col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, labels = names(mdat))

#=================================plot median income===============
sub.impute.dat$V1 <- seq.int(nrow(sub.impute.dat))
#sub.impute.dat[, lapply(.SD, FUN = mean.missing), .SDcols = "NY.GNP.PCAP.CD"]
missing <- which(is.na(sub.impute.dat$NY.GNP.PCAP.CD))
filter.missing <- sub.impute.dat[V1 %nin% missing ]
ggplot(filter.missing, aes(x = IncomeGroup, y = NY.GNP.PCAP.CD)) + 
  geom_boxplot() 

#=============================Other Plot=============================
M <- cor(after.imputation[, -c(1:6)], use = "complete.obs")
#corrplot(M, method = "pie")
qgraph(M)
corrplot(M, type="upper", order="hclust", tl.col="black", tl.srt=45)
hiv.dat <- after.imputation[, c("SH.DYN.AIDS.ZS", "NY.GNP.PCAP.CD")]
plot(hiv.dat$NY.GNP.PCAP.CD, hiv.dat$SH.DYN.AIDS.ZS)
```


